---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
load("www/join_demo.Rdata")
starwars.biodata <- starwars %>% select(name:eye_color)
starwars.species <- starwars %>% select(name,species)
starwars.home <- starwars %>% select(givenName=name,homeworld)
starwars.species <- starwars.species[-c(1,3,5),]
starwars.home <- starwars.home[-c(2,4,6),]
```

## Joining data sets

Often we will have different data sets that need to be combined.

For example we have three different data sets on Star Wars characters:

```{r dataview, exercise=TRUE}
head(starwars.biodata)
head(starwars.species)
head(starwars.home)
```

How do we combine these data sets?  We CANNOT just cbind them together because the names are not in the same order and not all data sets are complete.

We want to _join_ them by a common _key_.  In this case we can use the character's name as a key.

## Type of joins

The `dplyr` package in the `tidyverse` provides us with a number of join functions, four of which will be discussed here.

We choose a function based on how we want non-matching keys to be handled.

In each of these functions the first two arguments are the tables that we want to join and the third argument species which column(s) contain the keys

### Inner Join

`inner_join()` only keeps rows where the key is present in both data sets

```{r inner, exercise=TRUE}
starwars.bio.species <- inner_join(starwars.biodata,starwars.species, by = "name")
starwars.bio.species
```

### Left Join

`left_join()` keeps all rows from the left data set and keeps rows from the right data set where there is a key that matches the left.  Rows with non-matching keys in the right data set are discarded.

```{r left, exercise=TRUE}
starwars.bio.species <- left_join(starwars.biodata,starwars.species, by = "name")
starwars.bio.species
```

There is an analogous `right_join`

### Full join.

`full_join()` keeps all rows from left or right, regardless of whether or not there is a matching key.

```{r full, exercise=TRUE}
starwars.bio.species <- full_join(starwars.biodata,starwars.species, by = "name")
starwars.bio.species
```

## Specifying keys

What if the key columns have different names in the two data sets? you can specify the equivalency as shown here:

```{r by, exercise = TRUE}
starwars.species
starwars.home #oops different columns names for the key!

starwars.species.home <- left_join(starwars.species, starwars.home, by=c("name"="givenName"))

starwars.species.home
```

## Practice

We have three data sets containing information on _Brassica rapa_ genes that have been BLASTED agains _Arabidopsis thaliana_ to do a rough annotation.

* `gene.info` has information on the genomic location of the _B. rapa_ gene.
* `blast.info` has info on the best hit when the _B. rapa_ gene was BLASTED again _Arabidopsis_
* `At.info` has information on Arabidopsis gene function.

### Examine the data sets

First use `head` to take a look a each of the three data sets.
```{r head, exercise=TRUE}

```


### Practice 1

Create a new data frame that has information on the best BLAST hit for each Brassica gene.  Keep all of the Brassica genes in `gene.info` even if they don't have information in the `blast.info` data frame.  Discard any `blast.info` rows that are not matched in `gene.info`.

```{r practice1, exercise=TRUE}

```


### Practice 2

Repeat what you did above, but now do an additional join to add in the A.t. gene info.  Only keep rows where there is a matching entry in both data frame.

```{r practice2, exercise=TRUE}

```

## End

This has been a brief intro to joining.  It is a pretty complex topic, for more info, see [Relational Data in the R for Data Science book](http://r4ds.had.co.nz/relational-data.html)
