---
title: "Gather and Spread"
tutorial:
  version: 1.0
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
fruit <- read_csv("www/fruit_all.csv")
tomato <- read_csv("www/tomato_small.csv") %>% filter(!duplicated(paste0(species,trt)))
tomato.long <- tomato %>% gather(key="phenotype", value = "value", hyp:int4)
fruit.long <- fruit %>% gather(key="data_type", value = "value", -fruit)
```


## Introduction to Gather and Spread

### Wide data

There are two main ways that spreadsheets can be organized.  The __wide__ data format has one individual or sample per row and observations spread across multiple columns.

__Wide Data Example 1__  
There is one fruit type per row and multiple observations types across the columns.
```{r, echo=FALSE}
head(fruit)
```


__Wide Data Example 2__  
There is one tomato plant per row and multiple observations of different phenotypes in the columns:
```{r, echo=FALSE}
head(tomato)
```

### Long data

Alternatively, in the __long__ data format there is a single observation per row, and a key that indicates the type of observation.

__Long Data Example 1__
```{r}
fruit %>% gather(key = "observation_type", value = "value", -fruit) %>% arrange(fruit)
```

__Long Data Example 2__
```{r, echo=FALSE}
tomato %>% gather(key="phenotype", value = "value", hyp:int4)
```

### Who cares?

Typically data arrives in the wide format.  Often to accomplish the analyses you will want to do in R you will need it in the long format.  For example, ggplot only deals with a single column of values per aesthetic.  If you want to plot multiple categories of observations at once you need them gathered into the long format.  For example to make the graph below, the data has to be in the long format:

```{r, echo=FALSE}
tomato %>% gather(key="phenotype", value = "value", hyp:int4) %>%
  ggplot(aes(x=species, y=value, fill=phenotype)) + 
  geom_col() +
  facet_wrap(~trt) + 
  theme(axis.text.x = element_text(angle = 90))
```


## Gathering

To convert wide format to long format we use the `gather()` function from tidyverse.  It is appropriately named, because we are _gathering_ columns of data.

### Intro to gather

Let's look again at the original fruit data set:
```{r, echo=FALSE}
fruit
```

Here is how we gather it.
```{r gather1, exercise=TRUE}
fruit.long <- fruit %>% 
  gather(key="data_type", # the column name of the new column that will contain the key 
         value="value", # the column name of the column that will contain the observations
         color, shape, size # the column names that we want to gather.
  )
fruit.long
```

Pay attention to the arguments.

* `key` defines the column name of the new column that will contain the key 
* `value` defines the column name of the column that will contain the actual data
* The remaining arguments specify the columns that are to be gathered.

```{r gather-quiz1}
quiz(
  question("Where do the values in the `key` column come from?",
    answer("Row names"),
    answer("Data table values"),
    answer("Column names of the columns being gathered", correct = TRUE),
    answer("Column names of the columns NOT being gathered"),
    answer("R magic!")
  )
)
```

```{r gather-quiz2}
quiz(
  question("Where did the values from the `color`, `shape` and `size` columns go?",
    answer("Into the `key` column"),
    answer("Into the `value` column", correct = TRUE),
    answer("They vanished!")
  )
)
```

### Column specification

There are multiple ways that you can specify the columns to be gathered.

*You can list them all, as done in the first example:

```{r gather2, exercise=TRUE}
fruit.long <- fruit %>% 
  gather(key="data_type", # the column name of the new column that will contain the key 
         value="value", # the column name of the column that will contain the observations
         color, shape, size # the column names that we want to gather.
  )
fruit.long
```

*You can specify them as a range from the first to last column that you want:

```{r gather3, exercise=TRUE}
fruit.long <- fruit %>% 
  gather(key="data_type", # the column name of the new column that will contain the key 
         value="value", # the column name of the column that will contain the observations
         color:size # the column names that we want to gather.
  )
fruit.long
```

*You can instead use the `-` to indicate columns that you do not want to gather; all others will be included.  Try it by changing the code below

```{r gather4, exercise=TRUE}
fruit.long <- fruit %>% 
  gather(key="data_type", # the column name of the new column that will contain the key 
         value="value", # the column name of the column that will contain the observations
         -COLUMN_NAME_TO_EXCLUDE
  )
fruit.long
```

### Gather the tomato data

Practice what you learned above to gather the hyp, int1, int2, int3, and int4 columns together in the tomato data set.  Your end product should look like this:

```{r, echo=FALSE}
tomato %>% gather(key="phenotype", value = "value", hyp:int4)
```

```{r tomato-gather, exercise=TRUE}

```

## Spread your data

What if you need to go the other direction, from long format to wide?  For this we use the `spread()` function, to spread our values across multiple columns.

If we start with a long data set:

```{r, echo=FALSE}
fruit.long
```

We can spread it as follows:
```{r spread-example, exercise=TRUE}
fruit.long %>% 
  spread(key = "data_type", # which column holds the key?  
         value = "value") # which column contains the values?
```

Now try spreading the the tomato data to recreate the original format.  Modify the code below.

```{r spread-practice, exercise=TRUE}
tomato.long %>%
  spread(...)
```

## More information

More information on these functions is provided in [chapter 12 of R for Data science](http://r4ds.had.co.nz/tidy-data.html)

