---
title: "Introduction to Tidyverse"
tutorial:
  version: 1.0
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r, eval=FALSE, include=FALSE}
# this was used to set up the data sets.
library(babynames)
set.seed(99)
bnamesearly <- babynames %>% subset(year < 1901) %>% pull(name) %>% unique()
bnames10 <- sample(bnamesearly, 10)
bnames100 <- sample(bnamesearly, 100)
bnames1000 <- sample(bnamesearly, 1000)
bnamesall <- babynames$name %>% unique()
rev_words <- words %>% str_replace("^(.)(.*)(.)$", "\\3\\2\\1")
mywords <- words[words%in%rev_words]
mywords <- mywords[-1] %>% str_subset("^(.).*\\1$", negate = TRUE)
mywords <- mywords[c(1,2,3,6)]
rm(rev_words)
save.image(file="www/bnames.Rdata")
```


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
load("www/bnames.Rdata")
knitr::opts_chunk$set(echo = FALSE)
```


## Stringr and Regular Expressions

The goal of this tutorial is to introduce string manipulation and regular expression (wildcard) matching using the [stringr]() package.

The material covered here is also explained in greater detail in [Chapter 14 of the course textbook _R for Data Science_](https://r4ds.had.co.nz/strings.html) by Hadley Wickham and Garrett Grolemund.  I encourage you to use that book for clarification or greater depth.  It is free online.

stringr is part of the tidyverse and is loaded when you load tidyverse. It provides a set of functions with consistent syntax for manipulating strings.

### pre-requisites

To use the functions in this tutorial in your own R session you would need to use `library` to load the `tidyverse` package as shown below.  I've pre-loaded it for this tutorial.
```
library(tidyverse)
```

## Detect strings with `str_detect()`

The `str_detect()` function looks for the presence of a search pattern in a string and returns TRUE if it detects the pattern.

For example, consider the following 10 names:

```{r}
bnames10
```

Run the code below to search for names that contain "a":

```{r str-detect-1, exercise=TRUE}
bnames10 %>% str_detect("a")
```

`str_view()` performs a similar task but creates a visualization rather than returning output:

```{r str-detect-1view, exercise=TRUE}
bnames10 %>% str_view("a")
```

__Question:__
```{r str-detect-q, echo=FALSE}
question("Is str_detect() case sensitive?",
  answer("Yes", correct = TRUE),
  answer("No")
)
```

__Practice:__ Use str_detect to detect names with "A".
```{r str-detect-ex-2, exercise=TRUE}

```

## Subset strings with `str_subset()`

If we want to only retain those strings that match our pattern, we use `str_subset()`.

Run the code below to retain names that contain "a":

```{r str-subset-1, exercise=TRUE}
bnames10 %>% str_subset("a")
```

__Practice:__ keep names that contain "th"
```{r str-subset-ex-1, exercise=TRUE}

```

## Extract part of a string with `str_extract()`

If we want to extract the matching part of the string, we can use `str_extract()`

```{r str-extract-1, exercise=TRUE}
bnames10 %>% str_extract("a")
```

__Practice:__ Go ahead and try it, with the pattern of your choice.

```{r str-extract-2, exercise=TRUE}

```

This doesn't seem  very useful right now, but it will...

## Regular Expressions, part 1

Regular expressions are a powerful syntax for specifying wildcards and doing complex pattern matching.  They are a component of most (or all) modern computer languages.  We will only scratch the surface here, but this should be enough to do a fair number of basic tasks.

### The wildcard character.

The `.` character matches any character, so if we want to find names where there are two "a"s separated by any character, then: (note we are using a list of 1000 names now)

```{r period1, exercise=TRUE}
bnames1000 %>% str_subset("a.a")
```

Regular expressions can get pretty confusing. By adding the "match=TRUE" argument to`str_view()` we get a nice visualization that is similar to `str_subset()`:

```{r period1view, exercise=TRUE}
bnames1000 %>% str_view("a.a", match = TRUE)
```

__Now you try it:__ use `str_subset()` or `str_view()` on the the `bnames1000` list to show names where there are two "i"s separated by two characters

```{r period2, exercise=TRUE}

```

### Escaping characters

But what if you want to actually find a period? Just using `str_detect()`doesn't work because `.` matches everything:

```{r}
test <- c("has.a.period", "has_no_period", "hasNoPunctuation")
test %>% str_detect(".")
```

There are two options:

1) use `\\` to 'escape' it, that is to remove its special meaning.  Somewhat confusingly we have to use `\\` because we have to escape the `\` itself.

```{r}
test %>% str_detect("\\.")
```

2) If you do not need to use a regular expression at all in your search string you can embed it within `fixed()` which indicates that everything in the pattern should be used literally.

```{r}
test %>% str_detect(fixed("."))
```

### Specifying the number of matches.

The number of matching can be specified by appending special characters: (This is taken from the r help file on `regexp`)

* `*` specifies that the preceding item will be matched 0 or more times
* `?` the preceding item will be matched 0 or 1 times
* `+` The preceding item will be matched one or more times.
* `{n}` The preceding item is matched exactly n times.
* `{n,}` The preceding item is matched n or more times.
* `{n,m}` The preceding item is matched at least n times, but not more than m times.

__Practice:__ Start with the even larger `bnamesearly` and find all name where there is an "e" followed by 1 or 2 "z"s.  You results should look like this: 

```{r, echo=FALSE}
bnamesearly %>% str_subset("ez{1,2}")
```

```{r num-match, exercise=TRUE}
bnamesearly %>% str_subset("ez{1,2}")
```

## Regular Expressions, part 2

### Character sets

If we want to match any of a particular set of characters we can include the group in `[]`.

__Practice:__ Experiment with `[]` to find all names in the `bnames100` list that start with a vowel (you should find 19):

```{r brackets, exercise=TRUE}
bnames100 %>% str_subset("[]")
```

You can also specify ranges using "-" so that `[0-5]` matches numbers from 0 to 5 and `[j-n]` matches j, k, l, m, or n, `[A-Z]` is all capital letters, etc.  There are also some predefined classes but generally I do not recommend using them because the definitions change depending on your locale.

### Or

Somewhat related is the `|` character that serves as an or:

```{r jj1, exercise=TRUE}
bnamesall %>% str_subset("Jack|Jill") %>% head(15)
```

Parentheses are used to indicate the boundaries of the or statement.  Compare the following two code chunks:

No parentheses:
```{r jj2, exercise=TRUE}
bnamesall %>% str_subset("Jack|Jilly") 
```

With parentheses:
```{r jj3, exercise=TRUE}
bnamesall %>% str_subset("(Jack|Jill)y") 
```

Why are these last two different?  The first returns all names that contain "Jack" or "Jilly", whereas the second returns all names that contain "Jacky" or "Jilly".  

__Practice:__ The object `fruit` contains 80 fruit names.  

__1)__ Find all of those that contain "berry" or "pepper.

```{r bp1, exercise=TRUE}
fruit %>% str_subset("") 
```

__2)__ Find all fruit names that have a space " " preceding "berry" or "pepper"

```{r bp2, exercise=TRUE}
fruit %>% str_subset("") 
```

__3)__ Find all fruit names that contain "berry" or "pepper" but that do NOT have a preceding space 

```{r bp3, exercise=TRUE}
fruit %>% str_subset("") 
```

## Regular Expressions, part 3

Earlier you were asked to find all names that start with a vowel.  This was relatively easy because the first letter of the `bnames` list is capitalized. But what if you wanted all fruit that started with a vowel?  Luckily we can specify the beginning of a string with `^` and the end of a string with `$`.

__Practice:__ Use this information to find all fruit that start with a vowel:

```{r fruitvowel1, exercise=TRUE}
fruit %>% str_subset("") 
```

__Practice:__ Now try finding all fruit that start AND end with a vowel.

```{r fruitvowel2, exercise=TRUE}
fruit %>% str_subset("") 
```

__Practice:__ How about all fruit that start OR end with a vowel?

```{r fruitvowel3, exercise=TRUE}
fruit %>% str_subset("") 
```

## More `str_` functions

### Removing matches
`str_remove` removes the (first) match of the pattern and returns the rest of the string; `str_remove_all()` removes all occurrences of it..

To try this, try removing the first lowercase vowel and then all lowercase vowels from `bnames10`


```{r stremove, exercise=TRUE}
bnames10 %>% 
```

### Replacing matches
`str_replace()` and `str_replace_all()` will replace the match with a different string.  If we want to capitalize berry:

```{r strreplace1, exercise=TRUE}
fruit %>% str_replace("berry", "BERRY")
```

## Back references

You have already seen that parentheses can be used to clarify the order of search operations (e.g. "(Jack|Jill)y" ).  They can also define a group of matched characters that can be referred to subsequently in the search string or in a replace statement.  For example, if we want to find all names with pairs of characters:

```{r back1, exercise=TRUE}
fruit %>% str_view("(..)\\1", match = TRUE)
```

What is going on here?  the `..` matched any two characters.  Since `..` is enclosed in parentheses that defines a group and `\\1` refers back to the group.

__Practice:__  Try it!  Find all names in bnames100 that begin and end with the same character.  To make it simpler we will convert to lowercase first.

```{r back2, exercise=TRUE}
bnames100 %>% tolower() %>% 
```

You can define multiple match groups with separate sets of parentheses and back reference each one, the first group is `\\1`, the second one is `\\2` and so forth.  So if we wanted to swap the first and last letters of these words:

```{r}
mywords
```


```{r back3, exercise=TRUE}
str_replace(mywords, "^(.)(.*)(.)$", "\\3\\2\\1")
```

__Stop and Think:__ What does each term in the regular expression match?  How does this reverse the first and last characters?

## Mutate

All of these functions can be used with `mutate()` to transform columns of a dataframe or tibble.  Consider this list of names:

```{r, echo=FALSE}
names <- tibble(full_name=c("Maloof, Julin", "Davis, John", "Harmer, Stacey", "Harada, John", "Britt, Anne", "Brady, Siobhan", "Liu, Bo", "Sinha, Neelima", "Ledford, Joel"))
names
```

We can use `mutate()` to create a new column with just the first name:

```{r mut1, exercise=TRUE}
names %>%
  mutate(first_name=str_remove(full_name, ".*, "))
```

__Practice: __ Create a column with the last names.

```{r mut2, exercise=TRUE}
names %>%
  mutate(last_name=)
```

An alternative approach using `mutate()` and `str_replace()`:

```{r mut3, exercise=TRUE}
names %>%
  mutate(first_name=str_replace(full_name, ".*, (.*)", "\\1"))
```

__Stop and Think:__ In your own words explain the regular expressions above and how they work.

__Practice:__ Use `mutate()` and `str_replace()` to create a column with last names.

```{r mut4, exercise=TRUE}
names %>%
  mutate(last_name=)
```

__Practice:__ Use `mutate()` and `str_replace()` to create a column with the first name followed by the last name.  Your result should look like this:

```{r, echo=FALSE}
names %>%
  mutate(first_last=str_replace(full_name, "(.*), (.*)", "\\2 \\1"))
```

```{r mut5, exercise=TRUE}
names %>%
  mutate(last_name=)
```

## Resources and final words

Regular expression are very powerful.  They also take a lot of practice before they become intuitive.  The payoff is worth it.

A few helpful resources:

[stringr and regular expression cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf)

[Interactive visualization in R using regexplain](https://www.garrickadenbuie.com/project/regexplain/)

