---
title: "ggplot tutorial"
tutorial:
  version: 1.0
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(tidyverse)
library(learnr)
tomato <- read_csv("www/Tomato.csv")
tomatoH <- tomato %>% filter(trt=="H")
knitr::opts_chunk$set(echo = FALSE)
```


## ggplot

### ggplot intro

The [ggplot2](http://ggplot2.org) package is a package that allows you to rapidly make impressive plots from complex data sets.

In addition to this tutorial there are several excellent resources available on the web:

* The visualization chapter of [R for Data Science](http://r4ds.had.co.nz/data-visualisation.html)
* The Graphs section of [Cookbook for R](http://www.cookbook-r.com/Graphs/) (highly recommended! Especially good when you have a "how do I..." type of question)
* The official [ggplot documentation](http://ggplot2.tidyverse.org/reference/)

ggplot is part of the tidyverse, so let's load that library
```{r tidyverse, exercise = TRUE}
library(tidyverse)
```
(You can also just use `library(ggplot2)` if you don't want/need the other tidyverse packages)

### A first plot

We again work with the Tomato data set.  It has been preloaded for this tutorial.  To remind you, it has the following columns: 

* shelf, flat, col, row: information about where each plant was grown
* acs: the accession number for that strain
* trt: "H" is light with a high red:far\-red ratio; simulated sun.  "L" is light with a low red:far\- red ratio; simulated plant shade.
* days: days since germination
* date: date of measurement
* hyp: hypocotyl length (embryonic stem)
* int1-int4: individual internode lengths. Internodes are stem segments
* leafnum: which leaf was measured
* petleng: petiole length
* leafleng: leaf length
* leafwid: leaf width
* ndvi: a measure of plant vegetation density
* lat: latitude of origin
* long: longitude of origin
* alt: altitude of origin
* species: species of plant
* who: who measured the plants

I have also created a reduced data set, `tomatoH` that only contains measurements from the "H" treatment.
```{r head, exercise=TRUE}
head(tomatoH)
```

Let us begun by asking if there is a relationship between the length of the petiole (leaf stem) and the leaf of the leaf blade:

```{r firstplot, exercise=TRUE}
ggplot(data=tomato, 
       mapping = aes(x=petleng,y=leafleng)) +
  geom_point()
```

Yes, it looks like there is.  

Let's look at each line of the code above:

* `ggplot(data=tomato, ` ggplot is the main function and initiates the plot.  The data argument tells ggplot what data set to work on.
* `mapping = aes(x=petleng,y=leafleng) + ` Here we are telling ggplot which columns in our data set should be mapped to particular plot __aes__thetics.  The argument is called `mapping` and the input to mapping is the `aes()` function.  We use `aes()` to tell ggplot that petiole length should be mapped to the x axis and leaf length should be mapped to the y axis.
* notice that the line that beings with `mapping` ends with a `+` .  This tells R that we want to add to the plot created by the ggplot function.
* `geom_point()`.  Geoms (geometries) indicate what what type of plot we want to make.

We will look at these functions in more detail as the tutorial continues...

Note that since we loaded tidyverse we could also pipe our data into ggplot:

```{r pipe2ggplot, exercise=TRUE}
tomato %>% ggplot(mapping = aes(x=petleng,y=leafleng)) +
  geom_point()
```

## Aesthetics

In this section we will explore aesthetics.  As noted above, aesthetics control the relationship between your data and plot elements.

### color

One common aesthetic is the `color` aesthetic.  Change the code below so that color is mapped to the treatment (`trt`) column:

```{r color, exercise=TRUE}
tomato %>% ggplot(mapping = aes(x=petleng,y=leafleng)) +
  geom_point()
```

### shape

The `shape` aesthetic controls the shape of the plotted points.  

Create a plot of int3 vs int4 where color indicates trt, and shape indicates who measured the plant.

```{r shape-color, exercise=TRUE}

```





## Geoms

be sure to include multiple geoms...

## scales

## Facets

## Titles and Labels

## Saving plots 

ggsave

## Additional Resources



### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

